---
title: "Bayesian estimation with NIMBLE: nimbleLinkedSample()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian estimation with NIMBLE: nimbleLinkedSample()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## An experimental second linked backend

`nlmixr2bayes` links Stan to the compiled rxode2/nlmixr2est likelihood for
`est = "stan"` (see `vignette("nlmixr2bayes", "nlmixr2bayes")`). The same
idea -- sample against the *exact* compiled conditional likelihood every
other nlmixr2 method uses, rather than re-deriving the model in the
sampler's own language -- also works with
[NIMBLE](https://r-nimble.org/), through `nimbleLinkedSample()`:

```{r}
library(nlmixr2bayes)
fit <- nimbleLinkedSample(model, data)
```

**This is not (yet) a full `nlmixr2(..., est = "nimble")` method.**
`nimbleLinkedSample()` is a standalone function: it returns
`nimble::runMCMC()`'s raw samples matrix, not an `nlmixr2FitData` object --
no `$parFixed`, no IPRED/CWRES tables, no WAIC/LOO, no diagnostics beyond
what NIMBLE itself reports. Use it when you specifically want NIMBLE's
sampler (its own adaptive Metropolis-Hastings by default) against the same
linked likelihood the Stan backend uses; use `est = "stan"` for the
full estimation workflow.

### Current scope

Like `est = "stan"`, the model's ODE solving, residual-error model and
censoring stay in rxode2/nlmixr2est -- NIMBLE only ever sees a compiled
external call returning the conditional log-likelihood and its analytic
gradient (the gradient is computed but currently unused; see
[Why no HMC](#why-no-hmc) below). Within that shared design,
`nimbleLinkedSample()` covers a narrower model space than `est = "stan"`
today:

* **Diagonal omega only.** A correlated (non-diagonal) omega block is
  refused. Omega itself is fixed at its `ini()` value, not sampled --
  only the etas and thetas are.
* **No finite mixtures, no mu-referenced covariates, and no prior on an
  omega block.**
* **No `fix()`ed population parameters.** `literalFix = TRUE` (nlmixr2est's
  default and this function's) drops a fixed theta from the linked
  problem's parameter count entirely, so `nimbleLinkedSample()` refuses
  a model containing one up front, with a clear reason, rather than let a
  downstream count mismatch surface as a confusing error.
* **At least 2 free population parameters, and at least 2 total
  (subject, eta) combinations.** This is a NIMBLE limitation, not a
  choice: NIMBLE collapses a whole-array reference to a lower rank when
  its length is exactly 1, which breaks the custom distribution this
  function is built on. A model with exactly one free theta, or exactly
  one subject and one eta, is refused rather than silently mishandled.

Every one of these is enforced by `nimbleLinkedSample()` itself --
refusing loudly with a specific reason, not producing a plausible-looking
but wrong posterior.

## Priors and bounds

Free thetas' `ini({})` priors are honored using the *same*
`ini({})`/`prior()` syntax as `est = "stan"`, translated into NIMBLE's
distribution syntax and reused for its bounds:

```{r}
one.cmt <- function() {
  ini({
    tka <- 0.45
    tcl <- log(c(0, 2.7, 100))
    tv <- 3.45
    add.sd <- c(0, 0.7)
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    prior(tka) ~ dnorm(0.45, 1)
    prior(tcl) ~ dnorm(1, 1)
    prior(tv) ~ dnorm(3.45, 1)
    prior(add.sd) ~ dgamma(2, 1)
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}

samples <- nimbleLinkedSample(one.cmt, nlmixr2data::theo_sd,
                              niter = 4000L, nburnin = 2000L)
```

A theta's `ini()` bounds (support-promoted by its prior, exactly as in
`stanPriors()`) apply via NIMBLE's native truncation
(`T(dist, lower, upper)`); a theta with no explicit `prior()` falls back to
a weakly-informative normal at its own declared bounds, same as the
untruncated case above but without a user-specified distribution.

The supported prior catalogue is narrower than Stan's: `normal`,
`std_normal`, `lognormal`, `gamma`, `beta`, `exponential`, `uniform`,
`student_t`, and `logistic`. Each was individually checked against
NIMBLE's actual BUGS-registered distributions (not assumed from R's own
`dnorm`/`dgamma`/etc., which use different parameter conventions) --
notably `cauchy` is **not** supported: `dcauchy` is not natively usable in
NIMBLE model code at all (it would need a custom distribution registration
of its own), so a `cauchy` prior is refused with a clear message rather
than silently substituted or mishandled. An unsupported prior on a
population parameter refuses cleanly, naming the distribution.

## Why no HMC {#why-no-hmc}

`est = "stan"` samples with NUTS, using the linked likelihood's analytic
gradient. `nimbleLinkedSample()` uses NIMBLE's default samplers instead
(adaptive random-walk Metropolis, one per parameter block) -- entirely
derivative-free. The linked likelihood's gradient is still computed on
every call (the C entry point always returns it), but it is discarded:
NIMBLE's `nimbleExternalCall()` mechanism has no documented hook for
supplying a custom function's derivative to NIMBLE's own automatic
differentiation system, so plugging the analytic gradient into
`nimbleHMC`'s NUTS sampler is not currently possible without further
NIMBLE-side investigation. Expect worse mixing than the Stan backend on
strongly correlated or high-dimensional posteriors, particularly for the
etas.

## Reading the result

`nimbleLinkedSample()` returns exactly what `nimble::runMCMC()` returns: a
matrix of post-burn-in draws, columns named `theta[i]` (in `ini({})`
`ntheta` order) and `eta[subject, k]`.

```{r}
colnames(samples)
posterior_theta <- samples[, grepl("^theta", colnames(samples))]
colMeans(posterior_theta)
```

Everything in the `coda`/`posterior` ecosystem works directly on this
matrix (wrap it with `coda::as.mcmc()` or `posterior::as_draws_matrix()`
for trace plots, Rhat/ESS, etc.) -- `nimbleLinkedSample()` does not
compute any diagnostics itself.

## Verifying the link

`nimbleLinkedSample()` links to the exact same compiled
`nlmixr2bayes_cond_batch_theta` C entry point `est = "stan"` uses (through
a small standalone shim resolved via `R_GetCCallable` at runtime, compiled
on demand and cached) -- the same "one source of truth" conditional
likelihood, not a re-derivation. `nimbleBuildInfo()` probes the installed
NIMBLE's distribution registry and `nimbleExternalCall()` signature (no
compilation, milliseconds) and is checked automatically before the first
`nimbleLinkedSample()` call in a session, refusing clearly if a NIMBLE
version has changed something this backend depends on.
