---
title: "Bayesian estimation with est=\"stan\""
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian estimation with est="stan"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Automatic model generation: likelihood-level linking

`nlmixr2bayes` adds a Bayesian HMC estimation method to nlmixr2 that **automatically 
generates and compiles a Stan program**:

```{r}
library(nlmixr2bayes)
fit <- nlmixr2est::nlmixr2(model, data, est = "stan")
```

You write your model once (in nlmixr2 syntax), declare priors in the `ini({})` block, 
and let `est = "stan"` handle everything else: Stan program generation, compilation, 
linking, and sampling. **No Stan code to write** — this vignette is for you.

### When to use this approach

- ✅ You want Bayesian estimation of a complete nlmixr2 model
- ✅ You prefer automatic program generation over hand-coding Stan
- ✅ Your model fits the supported scope (ODE, `linCmt()`, normal/censored errors, finite mixtures)
- ✅ You want the same likelihood evaluated by every nlmixr2 method (FOCE, Stan, etc.)

### When to use hand-coded Stan instead

If you want **full control over your Stan program** — custom priors, derived quantities, 
multiple endpoints with mixed structures, or you're fitting **delay differential equations** — 
see `vignette("rxstan-handcoded", "nlmixr2bayes")` for the rxstan approach.

---

Stan is **linked** to the compiled rxode2/nlmixr2est likelihood rather than
the model being translated into Stan's language: rxode2 solves the ODEs
(dosing events, `addl`, infusions, steady state, delay equations), and
nlmixr2est evaluates the same conditional likelihood every other nlmixr2
method uses -- one source of truth, with analytic gradients handed to Stan
through `precomputed_gradients()`.  Stan samples theta, the omega
parameters, and the individual etas jointly; the eta prior is fully owned
by Stan (non-centred by default), so its autodiff composes the supplied
gradients through the Omega parameterization.  A model with **no** etas
(a population-only or single-subject fit) runs as "tier 0" through
nlmixr2est's nlm path instead -- one external scalar carrying the whole
data log-likelihood with its analytic gradient.

Beyond the core: finite mixtures (`mix()`, 2 components) marginalize with
the Stan `log_sum_exp` pattern and put per-subject membership posteriors
in `fit$env$mixProb`; fixed-variance etas (`eta ~ fix(v)`) render as
constant Cholesky factors with the eta still sampled; and an estimated
transform-both-sides lambda (Box-Cox/Yeo-Johnson, including on
`logitNorm`/`probitNorm` bases) is handled with the DV-transform Jacobian
emitted Stan-side.

## Priors come from `ini({})`

Priors are declared next to the parameters they describe, with the
lotri/rxode2 prior syntax -- the distribution catalogue is one-to-one with
Stan's (`lotri::lotriPriorDists()`):

```{r}
one.cmt <- function() {
  ini({
    tka <- 0.45
    tcl <- log(c(0, 2.7, 100))
    tv <- 3.45
    add.sd <- c(0, 0.7)
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    prior(tka) ~ dnorm(0.45, 1)
    prior(tcl) ~ dnorm(1, 1)
    prior(tv) ~ dnorm(3.45, 1)
    prior(add.sd) ~ dcauchy(0, 2.5)
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}

fit <- nlmixr2est::nlmixr2(one.cmt, nlmixr2data::theo_sd, est = "stan",
                           control = stanControl(seed = 42))
```

Notes on how priors are handled:

- A parameter's `ini()` bounds become Stan constraints; `add.sd <- c(0, 0.7)`
  with `prior(add.sd) ~ dcauchy(0, 2.5)` is the half-Cauchy idiom
  (`real<lower=0>` + `cauchy(0, 2.5)`).
- A prior whose support is narrower than the declared bounds tightens the
  declaration (support promotion); truncation `T[L,U]` is emitted only when
  the dropped normalizer would be parameter-dependent.
- Omega blocks: write `prior(eta.x) ~ invWishart(nu)` for an inverse-Wishart
  on the block, or leave the block without a prior to get the announced
  default -- `lkj_corr_cholesky(2)` on the correlation plus half-Cauchy
  priors on each SD.
- A model with **no** priors is refused, and the error prints the exact
  `prior()` lines to add.  Bayesian output from silently invented priors
  would look publishable and be wrong.
- Parameters that are `fix()`ed are dissolved before generation
  (`literalFix = TRUE`); a prior on a fixed parameter is an error.

## Controlling the run

```{r}
stanControl(
  chains = 4, iter = 2000, warmup = 1000, seed = 42,
  adapt_delta = 0.8, max_treedepth = 10, # Stan's own sampler defaults
  init = "ini",          # chains start at the ini() estimates (jittered)
  etaParam = "noncentered",  # or "centered"; identical posterior (gate G13)
  cores = 4,          # subject-parallel threads INSIDE each evaluation
  print = 100         # nlmixr2est-style iteration table cadence (0 = off)
)
```

Parallelism has two axes.  By default (unix) the chains run in FORKED
processes (`chainCores`, inheriting the `rxode2::getRxThreads()` budget
capped at the chain count): each fork receives a copy-on-write
duplicate of the linked likelihood, and the draws are bit-identical to
a sequential run.  With `chainCores = 1` the chains run sequentially
and each likelihood evaluation is subject-parallel instead (`cores`,
default `getRxThreads()`; CRAN's two-core policy is honored
automatically through `OMP_THREAD_LIMIT`).  The ODE solving is the
rate-limiting step in most pharmacometric models, so that is where the
cores go either way.

With sequential chains (`chainCores = 1`), during sampling the familiar
nlmixr2est iteration table prints every
`print`-th log-density evaluation: the natural-scale thetas, the current
omega variances/covariances (`om.<eta>`, `cov.<eta1>.<eta2>` -- Stan
samples omega, so these move), and `-2*sum(conditional log-lik)`.  The
printed rows are recorded on the fit as `$parHistData`.

## Other Stan inference algorithms: ADVI

`stanControl(algorithm=)` selects the inference algorithm that runs on the
same generated program and the same linked likelihood:

* `"NUTS"` (default) -- the No-U-Turn HMC sampler; full Bayes.
* `"meanfield"` -- ADVI with independent Gaussians on the unconstrained
  scale (Stan's `rstan::vb()`).
* `"fullrank"` -- ADVI with one full-covariance Gaussian.
* `"pathfinder"` -- multi-path Pathfinder (Zhang, Carpenter, Gelman &
  Vehtari 2022), implemented in nlmixr2bayes against the compiled model's
  exact log density and analytic gradient (rstan does not expose the
  Pathfinder service): L-BFGS trajectories from jittered starts, an
  ELBO-selected local Gaussian per path, and Pareto-smoothed importance
  resampling over the path mixture.  Needs the `loo` package.

```{r, eval=FALSE}
fit <- nlmixr2(one.cmt, theo_sd, est = "stan",
               control = stanControl(algorithm = "meanfield"))
# or the sugar methods, like "foce"/"focei" in nlmixr2est:
fit <- nlmixr2(one.cmt, theo_sd, est = "advi")        # meanfield ADVI
fit <- nlmixr2(one.cmt, theo_sd, est = "pathfinder")  # Pathfinder
```

Everything downstream is identical -- the variational draws flow through
the same posterior-to-fit machinery, so `$parFixed`, `$etaObf`, `$omega`,
`$posteriorSummary`, WAIC/LOO and the iteration print all work -- and
ADVI is typically 5-10x faster than NUTS on these models (~40 s vs
~260 s on `theo_sd`).  But ADVI is an *approximation*: it understates
tails and posterior correlations.  The Pareto-\(\hat k\) diagnostic of
the importance ratios replaces Rhat/ESS/divergences on the fit
(`$stanDiagnostics$khat`, also shown in the fit header): \(\hat k \le
0.7\) means the approximation supports importance correction; above that
the fit warns (via `onDiagnostic`) that the approximation is unreliable
and NUTS should be used.  Hierarchical PK posteriors routinely exceed it
-- treat ADVI as the fast first look and NUTS as the answer.

`run = FALSE` returns the generated Stan program and data without touching
rstan -- useful for inspection and for codegen tests:

```{r}
code <- nlmixr2est::nlmixr2(one.cmt, nlmixr2data::theo_sd, est = "stan",
                            control = stanControl(run = FALSE))
cat(code$code)
```

## Reading the fit

The result is a first-class `nlmixr2FitData`: `print(fit)`, `fit$parFixed`,
IPRED/CWRES tables, `plot(fit)` all work.  On top of that:

- `fit$posteriorSummary` -- exact posterior quantiles for every monitored
  parameter.  The `$parFixed` confidence intervals are a **Wald (Gaussian)
  approximation** built from the posterior covariance; for skewed
  posteriors (SDs, small omegas) prefer the exact quantiles here.
- `fit$etaObf` -- posterior-mean etas (with `OBJI` = -2 x mean conditional
  log-likelihood per subject).  The FOCEi empirical Bayes estimates
  computed for the comparability row are kept as `fit$etaObfFocei`.

### Working with the rstan object

The full `rstan::stanfit` is kept on the fit and retrieved with

```{r}
sf <- fit$stanfit          # the rstan::stanfit (also fit$env$stanfit)
```

Everything in the rstan/Bayesian ecosystem works from there:

```{r}
rstan::summary(sf)$summary            # full summary incl. per-eta rows
draws <- rstan::extract(sf)           # posterior draws by parameter
rstan::extract(sf, permuted = FALSE)  # iterations x chains x parameter
rstan::traceplot(sf, pars = c("tka", "tcl", "tv"))
rstan::get_sampler_params(sf, inc_warmup = FALSE)

# bayesplot / posterior work directly on the stanfit
bayesplot::mcmc_pairs(sf, pars = c("tka", "tcl", "tv"))
posterior::as_draws_df(sf)
```

Parameter naming inside the Stan program: `.` in nlmixr2 names becomes `_`
(`add.sd` is `add_sd`), the etas are the transformed parameter
`eta[subject, k]`, and the omega blocks appear as `omegaOut` (the
covariance in the model's eta order).  `fit$stanCode` and `fit$stanData`
hold the generated program and its data list, so any draw-level question
can be answered against the exact program that produced it.

### The objective-function rows

`fit$objDf` carries several rows on the deviance scale:

- **FOCEi** -- a plug-in frequentist objective evaluated at the posterior
  point estimate, for comparability with other nlmixr2 methods.  It
  double-uses the data and ignores the prior; it is a bridge to the
  likelihoodist world, not a Bayesian quantity.
- **WAIC / LOO** -- information criteria computed from the per-subject
  conditional log-likelihood draws, so they are **leave-one-subject-out**
  quantities (the subject is the exchangeable unit of a hierarchical
  model) conditional on the eta factorization.  The full `loo` objects are
  at `fit$env$loo` / `fit$env$waic`.

### Diagnostics

Sampler diagnostics run right after sampling and again at print time:
divergences, Rhat, bulk/tail ESS, treedepth saturation.  Each produces an
actionable message (which knob to turn); `onDiagnostic` selects
warn/error/message/none, and the headline numbers print above the parameter
table in `fit$extra`.  Do not use estimates from a run with divergences or
Rhat above 1.01 -- raise `adapt_delta` toward 0.99, or reconsider the
model.

## What is refused (and why)

Rather than produce a silently wrong posterior, `est="stan"` refuses:

- mixtures with more than 2 components (2-component `mix()` models are
  supported, with per-subject membership posteriors in
  `fit$env$mixProb`), IOV (until an upstream sensitivity fix,
  nlmixr2est#952), and discrete priors (every `ini({})` parameter is
  real-valued).

Estimated transform-both-sides lambda (Box-Cox/Yeo-Johnson) is supported:
the linked conditional supplies the transformed-scale density with an
exact `d/dlambda` sensitivity (nlmixr2est#949), and the generator emits
the DV-transform Jacobian on the Stan side --
`target += (lambda - 1) * sumLogJac`, a pure data statistic per endpoint
-- so the assembled target is the correct untransformed-scale density
with an exact lambda gradient (both gate-verified).

Mu-referenced covariates are supported whether subject-constant (an
analytic chain-rule identity) or time-varying (the coefficient is treated
as an estimated regressor through the forward-sensitivity model, as in the
other nlmixr2est methods).

Estimated dose-handling parameters (`alag`/`f`/`dur`/`rate`) are supported
with nlmixr2est >= the nlmixr2/nlmixr2est#946 fix (their derivatives carry
the jump condition at the dose event); with an older nlmixr2est they are
refused with an explanation.

## How it is validated

See the README's oracle tables.  In brief: every gradient is
finite-difference verified; a quadrature marginal over the linked
conditional reproduces `est="agq"`; censored and `ll()` conditionals tie to
hand-computed textbook densities; prior-only sampling recovers every
declared prior; the full posterior matches a hand-written native-Stan
implementation of the same model; simulate-and-recover covers the
generative truth; and the evaluation is bitwise deterministic (the property
NUTS's reversibility assumes), with the ODE retry machinery live.
