---
title: "Bayesian estimation with est=\"nuts\""
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian estimation with est="nuts"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
## Every fit below is run for real and cached under inst/cache with
## nlmixr2save's `:=` operator, so this page shows live output without
## re-sampling on every build.  See vignettes/cacheSetup.R.
source("cacheSetup.R")
bayesCache("nlmixr2bayes-")
```

## Automatic model generation: likelihood-level linking

`nlmixr2bayes` adds a Bayesian HMC estimation method to nlmixr2 that
**automatically generates and compiles a Stan program**.  You write the model
once, in nlmixr2 syntax, declare priors in the `ini({})` block, and
`est = "nuts"` handles everything else: Stan program generation, compilation,
linking, and sampling.  **No Stan code to write** -- this vignette is for you.

`est = "nuts"` names the ALGORITHM (Stan's No-U-Turn HMC sampler) rather than
the software tool -- it is sugar for `est = "stan"` with NUTS, which is
`stanControl()`'s default anyway, so the two are equivalent here.

### When to use this approach

- You want Bayesian estimation of a complete nlmixr2 model
- You prefer automatic program generation over hand-coding Stan
- Your model fits the supported scope (ODE, `linCmt()`, normal/censored errors, finite mixtures)
- You want the same likelihood evaluated by every nlmixr2 method (FOCE, Stan, etc.)

### When to use hand-coded Stan instead

If you want **full control over your Stan program** -- custom priors, derived quantities,
multiple endpoints with mixed structures, or you're fitting **delay differential equations** --
see `vignette("rxstan-handcoded", "nlmixr2bayes")` for the rxstan approach.

---

Stan is **linked** to the compiled rxode2/nlmixr2est likelihood rather than
the model being translated into Stan's language: rxode2 solves the ODEs
(dosing events, `addl`, infusions, steady state), and
nlmixr2est evaluates the same conditional likelihood every other nlmixr2
method uses -- one source of truth, with analytic gradients handed to Stan
through `precomputed_gradients()`.  Stan samples theta, the omega
parameters, and the individual etas jointly; the eta prior is fully owned
by Stan (non-centered by default), so its autodiff composes the supplied
gradients through the Omega parameterization.  A model with **no** etas
(a population-only or single-subject fit) runs as "tier 0" through
nlmixr2est's nlm path instead -- one external scalar carrying the whole
data log-likelihood with its analytic gradient.

Beyond the core: finite mixtures (`mix()`, 2 components) marginalize with
the Stan `log_sum_exp` pattern and put per-subject membership posteriors
in `fit$env$mixProb`; fixed-variance etas (`eta ~ fix(v)`) render as
constant Cholesky factors with the eta still sampled; and an estimated
transform-both-sides lambda (Box-Cox/Yeo-Johnson, including on
`logitNorm`/`probitNorm` bases) is handled with the DV-transform Jacobian
emitted Stan-side.

## Priors come from `ini({})`

Priors are declared next to the parameters they describe, with the
lotri/rxode2 prior syntax -- the distribution catalog is one-to-one with
Stan's (`lotri::lotriPriorDists()`).

The model below is the standard one-compartment oral theophylline model, with
between-subject variability on `ka`, `cl` and `v` and an additive residual
error, plus one `prior()` line per estimated parameter:

```{r model}
library(nlmixr2)
library(nlmixr2bayes)

one.cmt <- function() {
  ini({
    tka <- 0.45
    tcl <- log(c(0, 2.7, 100))
    tv <- 3.45
    add.sd <- c(0, 0.7)
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    prior(tka) ~ dnorm(0.45, 1)
    prior(tcl) ~ dnorm(1, 1)
    prior(tv) ~ dnorm(3.45, 1)
    prior(add.sd) ~ dcauchy(0, 2.5)
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}
```

Fitting it is one call.  The `:=` here is nlmixr2save's caching assignment
(see `vignettes/cacheSetup.R`): the first render runs the sampler for real and
writes the fit into `inst/cache`, and every later render loads it back, so the
numbers on this page are the ones the code produced.  Written with `<-`
instead, the line is an ordinary fit that runs every time:

```{r fit}
fit := nlmixr2est::nlmixr2(one.cmt, nlmixr2data::theo_sd, est = "nuts",
                           control = nutsControl(seed = 42, chains = 2,
                                                 iter = 1000))
```

A short run (2 chains of 1000, half of them warmup) keeps this page quick to
rebuild and its cached fit small; the defaults are 4 chains of 2000, which is
what you want for a real analysis.

The result prints as an ordinary nlmixr2 fit, with the sampler's headline
diagnostics in the header and the posterior summarized in the usual parameter
table:

```{r fitPrint}
print(fit)
```

Notes on how priors are handled:

- A parameter's `ini()` bounds become Stan constraints; `add.sd <- c(0, 0.7)`
  with `prior(add.sd) ~ dcauchy(0, 2.5)` is the half-Cauchy idiom
  (`real<lower=0>` + `cauchy(0, 2.5)`).
- A prior whose support is narrower than the declared bounds tightens the
  declaration (support promotion); truncation `T[L,U]` is emitted only when
  the dropped normalizer would be parameter-dependent.
- Omega blocks: write `prior(eta.x) ~ invWishart(nu)` for an inverse-Wishart
  on the block, or leave the block without a prior to get the announced
  default -- `lkj_corr_cholesky(2)` on the correlation plus half-Cauchy
  priors on each SD.
- A model with **no** priors is refused, and the error prints the exact
  `prior()` lines to add.  Bayesian output from silently invented priors
  would look publishable and be wrong.
- Parameters that are `fix()`ed are dissolved before generation
  (`literalFix = TRUE`); a prior on a fixed parameter is an error.

That refusal is worth seeing, because the error is also the fix -- it prints
the `prior()` lines to paste back into `ini({})`:

```{r noPriors, error = TRUE}
noPriors <- function() {
  ini({
    tka <- 0.45
    tcl <- log(c(0, 2.7, 100))
    tv <- 3.45
    add.sd <- c(0, 0.7)
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}
nlmixr2est::nlmixr2(noPriors, nlmixr2data::theo_sd, est = "nuts")
```

Those suggested priors are derived from the model's own `ini()` values and
bounds; they are a starting point to edit, not a recommendation.

## Controlling the run

`stanControl()` collects everything about the run.  Calling it on its own
returns the control object, which is also how you can check what a setting
resolved to before spending an hour on a fit:

```{r control}
stanControl(
  chains = 4, iter = 2000, warmup = 1000, seed = 42,
  adapt_delta = 0.8, max_treedepth = 10, # Stan's own sampler defaults
  init = "ini",          # chains start at the ini() estimates (jittered)
  etaParam = "noncentered",  # or "centered"; both target the identical posterior
  cores = 4,          # subject-parallel threads INSIDE each evaluation
  print = 100         # nlmixr2est-style iteration table cadence (0 = off)
)
```

Parallelism has two axes.  By default (unix) the chains run in FORKED
processes (`chainCores`, inheriting the `rxode2::getRxThreads()` budget
capped at the chain count): each fork receives a copy-on-write
duplicate of the linked likelihood, and the draws are bit-identical to
a sequential run.  With `chainCores = 1` the chains run sequentially
and each likelihood evaluation is subject-parallel instead (`cores`,
default `getRxThreads()`; CRAN's two-core policy is honored
automatically through `OMP_THREAD_LIMIT`).  The ODE solving is the
rate-limiting step in most pharmacometric models, so that is where the
cores go either way.

With sequential chains (`chainCores = 1`), during sampling the familiar
nlmixr2est iteration table prints every
`print`-th log-density evaluation: the natural-scale thetas, the current
omega variances/covariances (`om.<eta>`, `cov.<eta1>.<eta2>` -- Stan
samples omega, so these move), and `-2*sum(conditional log-lik)`.  The
printed rows are recorded on the fit as `$parHistData`.

## Other Stan inference algorithms: ADVI and Pathfinder

`stanControl(algorithm=)` selects the inference algorithm that runs on the
same generated program and the same linked likelihood:

* `"NUTS"` (default) -- the No-U-Turn HMC sampler; full Bayes.
* `"meanfield"` -- ADVI with independent Gaussians on the unconstrained
  scale (Stan's `rstan::vb()`).
* `"fullrank"` -- ADVI with one full-covariance Gaussian.
* `"pathfinder"` -- multi-path Pathfinder (Zhang, Carpenter, Gelman &
  Vehtari 2022), implemented in nlmixr2bayes against the compiled model's
  exact log density and analytic gradient (rstan does not expose the
  Pathfinder service): L-BFGS trajectories from jittered starts, an
  ELBO-selected local Gaussian per path, and Pareto-smoothed importance
  resampling over the path mixture.  Needs the `loo` package.

Each has a sugar `est=` name -- these select the ALGORITHM instead of the
software tool, the way `"foce"`/`"focei"` do in nlmixr2est:

```{r advi}
fitAdvi := nlmixr2est::nlmixr2(one.cmt, nlmixr2data::theo_sd, est = "advi",
                               control = adviControl(seed = 42))
fitPf := nlmixr2est::nlmixr2(one.cmt, nlmixr2data::theo_sd, est = "pathfinder",
                             control = pathfinderControl(seed = 42))
```

Each sugar method has its own control -- `nutsControl()`, `adviControl()`,
`pathfinderControl()` -- which is `stanControl()` with that algorithm
forced and everything else unchanged.  Passing one *is* the `est=`, the
same way `foceControl()` implies `est="foce"` in nlmixr2est, so these two
lines are the same fit:

```{r sugarEquivalent, eval = FALSE}
nlmixr2est::nlmixr2(one.cmt, nlmixr2data::theo_sd, est = "advi",
                    control = adviControl(seed = 42))
nlmixr2est::nlmixr2(one.cmt, nlmixr2data::theo_sd, adviControl(seed = 42))
```

A contradiction is an error rather than a silent override, and the message
names the call that would have meant it:

```{r contradiction, error = TRUE}
adviControl(algorithm = "pathfinder")
```

The fit also reports back the control you wrote -- `fit$control` on the NUTS
fit is a `nutsControl`, not a bare `stanControl`:

```{r controlClass}
class(fit$control)
class(fitAdvi$control)
```

Everything downstream is identical -- the variational draws flow through the
same posterior-to-fit machinery, so `$parFixed`, `$etaObf`, `$omega`,
`$posteriorSummary`, WAIC/LOO and the iteration print all work.  Lining the
three algorithms up on the same model shows both halves of the trade: the
population parameters agree closely, and the run times do not.

```{r compare}
cbind(NUTS = fit$parFixedDf$Estimate,
      ADVI = fitAdvi$parFixedDf$Estimate,
      Pathfinder = fitPf$parFixedDf$Estimate) |>
  `rownames<-`(rownames(fit$parFixedDf)) |>
  round(3)
```

But ADVI is an *approximation*: it understates tails and posterior
correlations.  The Pareto-\(\hat k\) diagnostic of the importance ratios
replaces Rhat/ESS/divergences on those fits (`$stanDiagnostics$khat`, also
shown in the fit header): \(\hat k \le 0.7\) means the approximation supports
importance correction; above that the fit warns (via `onDiagnostic`) that the
approximation is unreliable and NUTS should be used.

```{r khat}
fitAdvi$stanDiagnostics$khat
fitPf$stanDiagnostics$khat
```

Hierarchical PK posteriors routinely exceed that threshold -- treat ADVI and
Pathfinder as the fast first look and NUTS as the answer.

`run = FALSE` returns the generated Stan program and data without touching
rstan -- useful for inspection and for codegen tests.  This is the program the
fit above actually sampled:

```{r code, message = TRUE}
code <- nlmixr2est::nlmixr2(one.cmt, nlmixr2data::theo_sd, est = "nuts",
                            control = stanControl(run = FALSE))
cat(code$code)
```

The messages above the program are the announced omega defaults: this model
leaves its three eta blocks without a `prior()`, so each diagonal gets a
half-Cauchy scaled from its `ini()` variance, and a correlated block would
also get `lkj_corr_cholesky(2)`.  They are printed rather than assumed
silently, for the same reason a model with no priors at all is refused.

Two things to notice in it.  The `functions{}` block declares
`nlmixr2_cond_all2()` -- the external hook Stan calls for the conditional
log-likelihood and its analytic gradient -- and nothing about the ODE system
appears anywhere, because rxode2 owns that.  And the declared parameter names
read like the model (`tka`, `add_sd`, `sd_eta_ka`), not like an anonymous
translation of it.

## Reading the fit

The result is a first-class `nlmixr2FitData`: `print(fit)`, `fit$parFixed`,
IPRED/CWRES tables, `plot(fit)` all work.  On top of that:

- `fit$posteriorSummary` -- exact posterior quantiles for every monitored
  parameter.  The `$parFixed` confidence intervals are a **Wald (Gaussian)
  approximation** built from the posterior covariance; for skewed
  posteriors (SDs, small omegas) prefer the exact quantiles here.

```{r posteriorSummary}
fit$posteriorSummary
```

The `om.<eta>` rows are the omega variances on their natural scale, and
`cov.<eta1>.<eta2>` rows (absent here -- this model has no correlated block)
would carry the covariances.  Compare the `2.5%`/`97.5%` columns for
`om.eta.ka` with the symmetric Wald interval in `$parFixed` to see why the
exact quantiles are the ones to quote for a variance.

- `fit$etaObf` -- posterior-mean etas (with `OBJI` = -2 x mean conditional
  log-likelihood per subject).  The FOCEi empirical Bayes estimates
  computed for the comparability row are kept as `fit$etaObfFocei`.

```{r etaObf}
head(fit$etaObf)
```

### Working with the rstan object

The full `rstan::stanfit` is kept on the fit and retrieved with

```{r stanfit}
sf <- fit$stanfit          # the rstan::stanfit (also fit$env$stanfit)
sf
```

Everything in the rstan/Bayesian ecosystem works from there.  `rstan::summary()`
gives the full table including the per-subject eta rows the nlmixr2 summary
folds away:

```{r rstanSummary}
head(rstan::summary(sf)$summary)
```

`extract()` gives the draws themselves, permuted (a list, one element per
parameter) or as the raw iterations x chains x parameter array:

```{r extract}
draws <- rstan::extract(sf)
names(draws)
dim(draws$tka)                        # 1000 post-warmup draws, chains pooled
dim(rstan::extract(sf, permuted = FALSE))
```

The sampler's own per-draw diagnostics -- step size, tree depth, divergences --
come from `get_sampler_params()`:

```{r samplerParams}
head(rstan::get_sampler_params(sf, inc_warmup = FALSE)[[1]], 3)
```

And the plotting/posterior packages take the `stanfit` directly:

```{r plots, fig.height = 3, fig.width = 7}
rstan::traceplot(sf, pars = c("tka", "tcl", "tv"))
```

```{r pairs, fig.height = 5, fig.width = 6}
bayesplot::mcmc_pairs(sf, pars = c("tka", "tcl", "tv"))
```

```{r posterior}
posterior::as_draws_df(sf)
```

Parameter naming inside the Stan program: `.` in nlmixr2 names becomes `_`
(`add.sd` is `add_sd`), the etas are the transformed parameter
`eta[subject, k]`, and the omega blocks appear as `omegaOut` (the
covariance in the model's eta order).  `fit$stanCode` and `fit$stanData`
hold the generated program and its data list, so any draw-level question
can be answered against the exact program that produced it:

```{r stanData}
names(fit$stanData)
identical(fit$stanCode, code$code)
```

### The objective-function rows

`fit$objDf` carries several rows on the deviance scale:

```{r objDf}
fit$objDf
```

- **FOCEi** -- a plug-in frequentist objective evaluated at the posterior
  point estimate, for comparability with other nlmixr2 methods.  It
  double-uses the data and ignores the prior; it is a bridge to the
  likelihoodist world, not a Bayesian quantity.
- **WAIC / LOO** -- information criteria computed from the per-subject
  conditional log-likelihood draws, so they are **leave-one-subject-out**
  quantities (the subject is the exchangeable unit of a hierarchical
  model) conditional on the eta factorization.  The full `loo` objects are
  at `fit$loo` / `fit$waic` (also `fit$env$loo` / `fit$env$waic`), which is
  where the Pareto-k table and the standard errors live:

```{r loo}
fit$loo
```

With only 12 subjects, high Pareto-k values here are expected -- each subject
carries a large share of the information, which is exactly the situation
leave-one-out importance sampling struggles with.

### Diagnostics

Sampler diagnostics run right after sampling and again at print time:
divergences, Rhat, bulk/tail ESS, treedepth saturation.  Each produces an
actionable message (which knob to turn); `onDiagnostic` selects
warn/error/message/none, and the headline numbers print above the parameter
table in `fit$extra`.

```{r diagnostics}
fit$stanDiagnostics
```

Do not use estimates from a run with divergences or Rhat above 1.01 -- raise
`adapt_delta` toward 0.99, or reconsider the model.  `$messages` is empty when
every check passed.

## What is refused (and why)

Rather than produce a silently wrong posterior, `est="stan"` refuses
mixtures with more than 2 components and discrete priors (every
`ini({})` parameter is real-valued).  2-component `mix()` models are
supported, with per-subject membership posteriors in `fit$env$mixProb`.

A discrete prior is refused at generation time, before anything is compiled:

```{r discrete, error = TRUE}
discreteModel <- one.cmt |> ini(prior(tka) ~ dpois(2))
nlmixr2est::nlmixr2(discreteModel, nlmixr2data::theo_sd, est = "nuts",
                    control = stanControl(run = FALSE))
```

IOV is supported: `iov.x ~ v | OCC` expands into per-occasion
fixed-variance etas scaled by an estimated magnitude theta, the same way
nlmixr2est's own methods treat occasion-level variability.

Estimated transform-both-sides lambda (Box-Cox/Yeo-Johnson) is supported:
the linked conditional supplies the transformed-scale density with an
exact `d/dlambda` sensitivity, and the generator emits the DV-transform
Jacobian on the Stan side -- `target += (lambda - 1) * sumLogJac`, a pure
data statistic per endpoint -- so the assembled target is the correct
untransformed-scale density with an exact lambda gradient (both
FD-verified).

Mu-referenced covariates are supported whether subject-constant (an
analytic chain-rule identity) or time-varying (the coefficient is treated
as an estimated regressor through the forward-sensitivity model, as in the
other nlmixr2est methods).

Estimated dose-handling parameters (`alag`/`f`/`dur`/`rate`) are
supported (their derivatives carry the jump condition at the dose
event).

## How it is validated

See the README's oracle tables.  In brief: every gradient is
finite-difference verified; a quadrature marginal over the linked
conditional reproduces `est="agq"`; censored and `ll()` conditionals tie to
hand-computed textbook densities; prior-only sampling recovers every
declared prior; the full posterior matches a hand-written native-Stan
implementation of the same model; simulate-and-recover covers the
generative truth; and the evaluation is bitwise deterministic (the property
NUTS's reversibility assumes), with the ODE retry machinery live.
