---
title: "Hand-coded Stan models with rxstan backend"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hand-coded Stan models with rxstan backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Hand-coded Stan: ODE-level linking

While `nlmixr2(model, data, est = "stan")` automatically generates a Stan program
(see `vignette("nlmixr2bayes", "nlmixr2bayes")`), you may want to **write your own Stan code**
to have full control over priors, derived quantities, and model structure.

The `rxstan` component of `nlmixr2bayes` lets you do exactly that. You register an rxode2 model
as a solver backend via `rxsRegister()`, then write ordinary Stan that declares and calls it.

### When to use this approach

- ✅ You want **full control over your Stan program**
- ✅ You need **delay differential equations** (Stan has no DDE solver)
- ✅ You want to write **custom priors or derived quantities** in Stan
- ✅ Your model has **complex structure** beyond what automatic generation supports
- ✅ You're fitting **hand-coded rxode2 models** outside the nlmixr2 framework

### When to use automatic generation instead

If you have a complete **nlmixr2-style model** with priors in the `ini({})` block, 
prefer `nlmixr2(model, data, est = "stan")` — it's simpler, keeps one source of truth 
for the likelihood, and is validated against every other nlmixr2 method. 
See `vignette("nlmixr2bayes", "nlmixr2bayes")`.

---

## What this does

While `nlmixr2(model, data, est = "stan")` automatically generates and compiles a Stan program,
you may want to write your own Stan code to have full control over priors, derived quantities,
and model structure. The `rxstan` component of `nlmixr2bayes` lets you do exactly that.

Stan can already solve ODEs, but it does so by autodiffing *through* the integrator, which is 
expensive, and its interface has no concept of a dose, a lag time, or a delay differential equation.

rxode2 has all of that, and it computes analytic first-order sensitivities $\partial y / \partial \theta$ 
by augmenting the system with the forward sensitivity equations. Those are exactly the numbers Stan's 
reverse-mode tape wants.

So rxstan hands the solve to rxode2 and pushes rxode2's derivatives onto Stan's tape directly, 
through `precomputed_gradients()`. Stan never differentiates the solver; it treats `rx_solve()` 
as a function whose partials are simply known.

```
Stan  --- theta -->  rxstan bridge (C++)  --- theta -->  rxode2
      <-- y, dy/dtheta ---                <-- states + rx__sens_* ---
```

On a one-compartment oral model with 100 observations, per gradient:

| solver             | time per gradient |
|--------------------|-------------------|
| rxstan             | 0.055 ms          |
| Stan `ode_rk45`    | 0.125 ms          |
| Stan `ode_bdf`     | 0.400 ms          |

Speed is the smaller half of the argument. The larger half is that dosing events and DDEs 
are not expressible in Stan's ODE interface at all.

`rstan` is required rather than optional: rxode2 has to be callable from the same process, 
so the model must be compiled and run in-process.

## The quick path: generate the Stan program

If you already have an nlmixr2-style model, `rxsStanFromUi()` writes the Stan program, 
registers the solver, and assembles the `standata`.

```{r model, eval=FALSE}
library(nlmixr2)

oneCmt <- function() {
  ini({
    tka <- 0.5
    tcl <- 1.0
    tv <- 3.4
    eta.cl ~ 0.09
    add.sd <- 0.2
  })
  model({
    ka <- exp(tka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv)
    d/dt(depot) <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}
```

Some data. Any data frame with `id`, `time`, `dv` and the usual event columns will do.

```{r data, eval=FALSE}
## sampling has to cover the absorption phase, or ka is not identified
times <- c(0.25, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 12, 24)
dat <- do.call(rbind, lapply(1:8, function(i) {
  e <- rxode2::et(amt = 100, cmt = "depot")
  e <- rxode2::et(e, times)
  d <- as.data.frame(e)
  d$id <- i
  d$dv <- 0
  d
}))
```

Generate the Stan program and solver handle:

```{r gen, eval=FALSE}
gen <- rxsStanFromUi(oneCmt, dat)
names(gen)
cat(gen$code)
```

`gen$code` is an ordinary Stan program, `gen$handle` is the live solver, and `gen$standata` 
is what you pass to `sampling()`. Note that `cp = center / v` is formed **in Stan**, not in 
rxode2 — see "What crosses the boundary" below.

## Writing your own Stan program

`rxsStanFromUi()` is convenient for quick fits, but to write your own program, use `rxsRegister()` 
to set up the solver handle and build the Stan code yourself.

```stan
functions {
  // Declared but not defined here; --allow_undefined tells the compiler
  // to look for it in the linked C++ bridge
  vector rx_solve(int handle, vector p);
}
data {
  int<lower=1> nObs;
  int<lower=1> handle;
  vector<lower=0>[nObs] cpObs;
}
parameters {
  real lka;
  real lcl;
  real lv;
  real<lower=0> sigma;
}
model {
  vector[3] p = [lka, lcl, lv]';
  vector[nObs] centre = rx_solve(handle, p);
  vector[nObs] cp = centre / exp(lv);

  // Your priors here
  lka ~ normal(0, 1);
  lcl ~ normal(1.4, 1);
  lv ~ normal(3.4, 1);
  sigma ~ normal(0, 1);

  // Likelihood
  log(cpObs) ~ normal(log(cp), sigma);
  target += -sum(log(cpObs));  // Jacobian of the log transform
}
```

From R:

```{r handcoded, eval=FALSE}
library(nlmixr2)

oneCmt <- function() {
  ini({
    tka <- 0.5
    tcl <- 1.0
    tv <- 3.4
    eta.cl ~ 0.09
    add.sd <- 0.2
  })
  model({
    ka <- exp(tka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv)
    d/dt(depot) <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}

# Register the model and set up the solver
h <- rxsRegister(oneCmt, events = dat, 
                 sens = c("lka", "lcl", "lv"), 
                 output = "center")

# Compile the Stan program
f <- tempfile(fileext = ".stan")
writeLines(stan_code, f)
sm <- rxsStanModel(f, modelName = "one_cmt")

# Prepare data for Stan
standata <- list(
  nObs = nrow(dat[is.na(dat$evid) | dat$evid == 0, ]),
  handle = h,
  cpObs = dat$dv[is.na(dat$evid) | dat$evid == 0]
)

# Sample
fit <- rstan::sampling(sm, data = standata, 
                       init = rxsInit(h), 
                       chains = 4, iter = 2000)
```

## What crosses the boundary

ODE **states**, never derived `lhs` quantities. rxode2 only emits `rx__sens_<state>_BY_<param>__` 
for states, so any derived prediction (a concentration `central / v`, say) must be formed in the 
Stan program, where Stan differentiates it itself. That keeps exactly one differentiation boundary.

In the example above, `depot` and `center` are states; `cp` is derived and formed in Stan.

## Checking the gradient

Before fitting, use `rxsCheckGradient()` to verify that the analytic sensitivities agree with 
finite differences:

```{r check, eval=FALSE}
# After compiling the Stan model
fit_init <- rstan::sampling(sm, data = standata, chains = 0)

# Test parameter vector
test_pars <- c(lka = 0.5, lcl = 1.0, lv = 3.4, sigma = 0.2)

# Check gradient
check <- rxsCheckGradient(fit_init, test_pars, h = 1e-5)
print(check)
```

Look for relative differences close to 1e-6 or better. Anything much worse suggests 
a problem with the bridge or the solver tolerances.

## Limits

- **rstan only** — rxode2 must be in-process.
- **Single-threaded** — do not use the bridge inside `map_rect()` or `reduce_sum()`. 
  rxode2's solve structure is a process singleton, so concurrent entry is detected and throws 
  rather than corrupting silently. Run chains as separate processes; leave subject-level 
  parallelism to rxode2.
- **First-order only** — `precomputed_gradients()` is `var`-only, so `laplace_marginal` 
  and RHMC are out.
