---
title: "Hand-coded Stan models with rxstan backend"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hand-coded Stan models with rxstan backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
## This vignette compiles one real Stan program (about a minute) and samples
## it; the fit itself takes seconds, so nothing here is cached.
```

## Hand-coded Stan: ODE-level linking

While `nlmixr2(model, data, est = "nuts")` automatically generates a Stan program
(see `vignette("nlmixr2bayes", "nlmixr2bayes")`), you may want to **write your own Stan code**
to have full control over priors, derived quantities, and model structure.

The `rxstan` component of `nlmixr2bayes` lets you do exactly that. You register an rxode2 model
as a solver backend via `rxsRegister()`, then write ordinary Stan that declares and calls it.

### When to use this approach

- You want **full control over your Stan program**
- You need **delay differential equations** (Stan has no DDE solver -- see
  `vignette("rxstan-dde", "nlmixr2bayes")`)
- You want to write **custom priors or derived quantities** in Stan
- Your model has **complex structure** beyond what automatic generation supports
- You're fitting **hand-coded rxode2 models** outside the nlmixr2 framework

### When to use automatic generation instead

If you have a complete **nlmixr2-style model** with priors in the `ini({})` block,
prefer `nlmixr2(model, data, est = "nuts")` -- it's simpler, keeps one source of truth
for the likelihood, and is validated against every other nlmixr2 method.
See `vignette("nlmixr2bayes", "nlmixr2bayes")`.

---

## Why link instead of translate

Stan can already solve ODEs, but it does so by autodiffing *through* the integrator, which is
expensive, and its interface has no concept of a dose, a lag time, or a delay differential equation.

rxode2 has all of that, and it computes analytic first-order sensitivities $\partial y / \partial \theta$
by augmenting the system with the forward sensitivity equations. Those are exactly the numbers Stan's
reverse-mode tape wants.

So rxstan hands the solve to rxode2 and pushes rxode2's derivatives onto Stan's tape directly,
through `precomputed_gradients()`. Stan never differentiates the solver; it treats `rx_solve()`
as a function whose partials are simply known.

```
Stan  --- theta -->  rxstan bridge (C++)  --- theta -->  rxode2
      <-- y, dy/dtheta ---                <-- states + rx__sens_* ---
```

On a one-compartment oral model with 100 observations, per gradient:

| solver             | time per gradient |
|--------------------|-------------------|
| rxstan             | 0.055 ms          |
| Stan `ode_rk45`    | 0.125 ms          |
| Stan `ode_bdf`     | 0.400 ms          |

Speed is the smaller half of the argument. The larger half is that dosing events and DDEs 
are not expressible in Stan's ODE interface at all.

`rstan` is required rather than optional: rxode2 has to be callable from the same process, 
so the model must be compiled and run in-process.

## The quick path: generate the Stan program

If you already have an nlmixr2-style model, `rxsStanFromUi()` writes the Stan program,
registers the solver, and assembles the `standata` -- no Stan to write, but the program is
yours to edit afterwards.

```{r libs}
library(nlmixr2)
library(nlmixr2bayes)
```

```{r model}
oneCmt <- function() {
  ini({
    tka <- 0.5
    tcl <- 1.0
    tv <- 3.4
    eta.cl ~ 0.09
    add.sd <- 0.2
  })
  model({
    ka <- exp(tka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv)
    d/dt(depot) <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}
```

Some data. Any data frame with `id`, `time`, `dv` and the usual event columns will do;
here it is simulated from the model itself so the numbers below have something real to
recover.  The sampling times deliberately cover the absorption phase -- without early
points `ka` is not identified.

```{r data}
set.seed(42)
times <- c(0.25, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 12, 24)
ev <- rxode2::et(amt = 100, cmt = "depot")
ev <- rxode2::et(ev, times)
ev <- rxode2::et(ev, id = 1:8)

sim <- rxode2::rxSolve(oneCmt, ev, returnType = "data.frame")
dat <- as.data.frame(ev)
names(dat) <- tolower(names(dat))
dat$dv <- NA_real_
dat$dv[dat$evid == 0] <- sim$sim   # dose rows carry no observation
head(dat, 3)
```

Generate the Stan program and solver handle:

```{r gen}
gen <- rxsStanFromUi(oneCmt, dat)
names(gen)
```

`gen$code` is an ordinary Stan program, `gen$handle` is the live solver, and `gen$standata`
is what you pass to `sampling()`:

```{r genCode}
cat(gen$code)
```

Note that `cp = center / v` is formed **in Stan**, in `transformed parameters`, not in
rxode2 -- see "What crosses the boundary" below.  The `p` vector packs each subject's
thetas and etas into one block-major parameter vector, which is what `perSubject = TRUE`
on the handle expects.

`gen$standata` is the data list, and `rxsInit()` turns the model's `ini()` values into a
Stan `init` function -- one starting point per chain, with the fixed effects jittered so
`Rhat` stays honest:

```{r genData}
str(gen$standata)
str(rxsInit(gen)(1))
```

Compiling and sampling it is then the ordinary rstan pair, exactly as in the hand-written
case below:

```{r genFit, eval = FALSE}
sm <- rxsStanModel(tempfile(fileext = ".stan"))  # after writing gen$code to it
fit <- rstan::sampling(sm, data = gen$standata, init = rxsInit(gen),
                       chains = 4, iter = 2000)
```

## Writing your own Stan program

`rxsStanFromUi()` is convenient for quick fits, but to write your own program, use
`rxsRegister()` to set up the solver handle and build the Stan code yourself.

Start from a plain rxode2 model, written on the log scale because that is how Stan will
parameterize it.  `cp = center / v` is deliberately *not* part of it -- only ODE states
cross the bridge:

```{r pkModel}
pkModel <- "
ka  <- exp(lka)
cl  <- exp(lcl)
v   <- exp(lv)
d/dt(depot)  <- -ka * depot
d/dt(center) <-  ka * depot - (cl / v) * center
"
```

Here is the Stan program that drives it.  It is the one shipped with the package as
`system.file("stan", "pk_1cmt_oral.stan", package = "nlmixr2bayes")`, so the code compiled
below is literally this text:

```stan
functions {
  // Declared but not defined here; --allow_undefined tells the compiler
  // to look for it in the linked C++ bridge
  vector rx_solve(int handle, vector p);
}
data {
  int<lower=1> nObs;
  int<lower=1> handle;
  vector<lower=0>[nObs] cpObs;
}
parameters {
  real lka;
  real lcl;
  real lv;
  real<lower=0> sigma;
}
model {
  vector[3] p = [lka, lcl, lv]';
  vector[nObs] center = rx_solve(handle, p);
  vector[nObs] cp = center / exp(lv);

  // Your priors here
  lka ~ normal(0, 1);
  lcl ~ normal(1.4, 1);
  lv ~ normal(3.4, 1);
  sigma ~ normal(0, 1);

  // Likelihood
  log(cpObs) ~ normal(log(cp), sigma);
  target += -sum(log(cpObs));  // Jacobian of the log transform
}
```

Three things make it work: `rx_solve()` is *declared* and never defined (the bridge
supplies it), `handle` arrives as ordinary integer data, and the returned `center` is a
state -- `cp` is formed from it in Stan.

### Register the solver

`rxsRegister()` compiles the model with forward sensitivities with respect to `sens`, in
the exact order the Stan program packs them into `p`, and returns the integer handle:

```{r register}
truth <- c(lka = log(1.1), lcl = log(4), lv = log(30))

ev1 <- rxode2::et(amt = 100, cmt = "depot")
ev1 <- rxode2::et(ev1, c(0.25, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 12, 16, 24))

h <- rxsRegister(pkModel, events = ev1,
                 sens = c("lka", "lcl", "lv"),
                 output = "center",
                 atol = 1e-10, rtol = 1e-10)
h
```

The tolerances are deliberately tighter than an ordinary point-estimate run would need:
HMC's leapfrog integrator assumes the target is a smooth, reversible function of the
parameters, so solver error has to sit well below the gradient signal.

`rxsSolve()` calls the same C entry point Stan does, which makes it the quickest way to
see what actually crosses the boundary -- column 1 is the state, the rest is its Jacobian:

```{r solve}
sol <- rxsSolve(h, unname(truth))
head(sol, 3)
```

Simulated observations, with a 15% lognormal residual to match the program's
`log(cpObs) ~ normal(log(cp), sigma)`:

```{r simData}
set.seed(42)
cp <- sol[, 1] / exp(truth[["lv"]])
cpObs <- cp * exp(stats::rnorm(length(cp), 0, 0.15))
standata <- list(nObs = length(cpObs), handle = h, cpObs = cpObs)
```

### Compile and fit

```{r compile}
sm <- rxsStanModel(system.file("stan", "pk_1cmt_oral.stan",
                               package = "nlmixr2bayes"),
                   modelName = "pk_1cmt_oral")
```

`rxsStanModel()` is a thin wrapper over `rstan::stanc()` + `rstan::stan_model()` that
splices the bridge header in at global scope and keeps the model name unobfuscated, so
`rx_solve()` resolves.  It is the slow step of this vignette -- about a minute -- and the
only one.

```{r sample}
fit <- rstan::sampling(sm, data = standata, chains = 2, iter = 1000,
                       seed = 42, refresh = 0,
                       init = function(chain_id = 1) {
                         list(lka = 0.1, lcl = 1.4, lv = 3.4, sigma = 0.2)
                       })
print(fit, pars = c("lka", "lcl", "lv", "sigma"))
```

Sampling itself takes a couple of seconds: the whole posterior is 1000 draws over four
parameters, each gradient one rxode2 solve of the sensitivity-augmented system.  The
posterior means sit on the simulated truth:

```{r recover}
round(rbind(truth = truth,
            posterior = colMeans(as.matrix(fit, pars = c("lka", "lcl", "lv")))), 3)
```

`rxsSolveStats()` counts what the sampler actually asked the bridge for, including how
many proposals rxode2 refused (HMC proposes bad parameters routinely; each one becomes a
rejected proposal rather than an error):

```{r stats}
rxsSolveStats(h)
```

## What crosses the boundary

ODE **states**, never derived `lhs` quantities. rxode2 only emits `rx__sens_<state>_BY_<param>__` 
for states, so any derived prediction (a concentration `central / v`, say) must be formed in the 
Stan program, where Stan differentiates it itself. That keeps exactly one differentiation boundary.

In the example above, `depot` and `center` are states; `cp` is derived and formed in Stan.

## Checking the gradient

Before trusting a fit, use `rxsCheckGradient()` to verify that the analytic sensitivities
agree with finite differences of the model's own log density.  It needs a `stanfit` from
the compiled model -- the one just fitted will do, and `sampling(..., chains = 0)` is
enough if you have not fitted yet -- plus the parameter vector to test at, on the
*unconstrained* scale Stan works in:

```{r check}
rxsCheckGradient(fit, c(lka = 0.1, lcl = 1.4, lv = 3.4, sigma = 0.2))
```

Look for relative differences close to `1e-6` or better. Anything much worse suggests a
problem with the bridge or, far more often, with the solver tolerances -- the value and
the gradient come from two *different* integrations (the state equations and the
sensitivity equations), so they agree only as far as `atol`/`rtol` make them.  Re-run this
whenever you loosen them from the `rxsRegister()` defaults.

## Cleaning up

A handle holds a compiled rxode2 model and its solve structure for the life of the
session.  Release it when you are done, or `rxsReleaseAll()` for all of them:

```{r release}
rxsRelease(h)
rxsRelease(gen$handle)
rxsHandles()
```

## Limits

- **rstan only** -- rxode2 must be in-process.
- **Single-threaded** -- do not use the bridge inside `map_rect()` or `reduce_sum()`. 
  rxode2's solve structure is a process singleton, so concurrent entry is detected and throws 
  rather than corrupting silently. Run chains as separate processes; leave subject-level 
  parallelism to rxode2.
- **First-order only** -- `precomputed_gradients()` is `var`-only, so `laplace_marginal`
  and RHMC are out.

## See also

- `vignette("nlmixr2bayes", "nlmixr2bayes")` -- the automatic, likelihood-level
  approach; simpler when a complete nlmixr2 model with `ini({})` priors is
  all you need.
- `vignette("rxstan-dde", "nlmixr2bayes")` -- a full delay differential
  equation walkthrough using this same bridge.
- `vignette("torsten-comparison", "nlmixr2bayes")` -- how this bridge
  compares to hand-writing the same model against Torsten's Stan functions.
