---
title: "Delay differential equations through rxstan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Delay differential equations through rxstan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
## The est="nuts" fit below is run for real and cached under inst/cache with
## nlmixr2save's `:=` operator; the bridge-level material is a few seconds of
## rxode2 solving and is not cached.  See vignettes/cacheSetup.R.
source("cacheSetup.R")
bayesCache("dde-")
```

## Delay differential equations: a Stan limitation

**Stan has no DDE solver.** Not a slow one -- none. There is no `ode_*` function that accepts 
a lagged state, and you cannot write one in the Stan language, because the solution has to be 
interpolated at $t - \tau$ while it is still being computed.

However, **rxode2 has `delay()`**, and it propagates sensitivities through the delay
correctly.  Because nlmixr2bayes links Stan to a *likelihood* rather than translating a
model into the Stan language, that limitation never reaches Stan: whatever rxode2 can
solve, Stan can sample.

So a DDE is reachable **both** ways nlmixr2bayes offers, and this vignette walks the same
example through each:

1. **Automatically** -- `nlmixr2(model, data, est = "nuts")` on an ordinary nlmixr2 model
   containing `delay()`.  No Stan code, and nothing special to configure.
2. **Hand-coded** -- register the DDE as a solver backend with `rxsRegister()` and call it
   from a Stan program you wrote yourself (see
   `vignette("rxstan-handcoded", "nlmixr2bayes")` for the general pattern), which is also
   the level at which you can inspect the delayed sensitivities directly.

---

## A delayed indirect response

Turnover with delayed feedback: the loss term depends on the response one delay ago 
rather than now.

$$\frac{dR}{dt} = k_{in} - k_{out}\,R(t - \tau), \qquad R(s) = 1 \ \text{for}\ s \le 0$$

`R(0) <- 1` does double duty: it is the initial condition *and* the pre-history, so
`delay(R, tau)` returns 1 whenever $t - \tau \le 0$.  Everything is parameterized on the
log scale (`lkin`, `lkout`, `ltau`), which is the scale Stan samples on and keeps all
three positive without a constraint.

## Part 1: the automatic path

Written as an nlmixr2 model with priors, the delay is just another line -- there is
nothing DDE-specific in the `ini({})` block, and `tau` is estimated like any other theta:

```{r autoModel}
library(nlmixr2)
library(nlmixr2bayes)

ddeMod <- function() {
  ini({
    lkin <- log(0.7)
    lkout <- log(0.4)
    ltau <- log(1.5)
    add.sd <- c(0, 0.3)
    eta.kin ~ 0.05
    prior(lkin) ~ dnorm(log(0.7), 0.5)
    prior(lkout) ~ dnorm(log(0.4), 0.5)
    prior(ltau) ~ dnorm(log(1.5), 0.5)
    prior(add.sd) ~ dcauchy(0, 1)
  })
  model({
    kin <- exp(lkin + eta.kin)
    kout <- exp(lkout)
    tau <- exp(ltau)
    R(0) <- 1
    d/dt(R) <- kin - kout * delay(R, tau)
    R ~ add(add.sd)
  })
}
```

Six subjects, simulated from the model itself so there is a known truth to recover:

```{r autoData}
set.seed(42)
tmsAuto <- seq(0.5, 20, by = 1.5)
evAuto <- rxode2::et(tmsAuto)
evAuto <- rxode2::et(evAuto, id = 1:6)

simAuto <- rxode2::rxSolve(ddeMod, evAuto, returnType = "data.frame")
datAuto <- as.data.frame(evAuto)
names(datAuto) <- tolower(names(datAuto))
datAuto$dv <- simAuto$sim
head(datAuto, 3)
```

And that is the whole fit -- an ordinary `est = "nuts"` call.  `:=` is nlmixr2save's
caching assignment, so it runs once and is reloaded on later builds:

```{r autoFit}
fitAuto := nlmixr2est::nlmixr2(ddeMod, datAuto, est = "nuts",
                               control = nutsControl(seed = 42, chains = 2,
                                                     iter = 1000,
                                                     max_treedepth = 12L))
print(fitAuto)
```

The posterior sits on the simulating values:

```{r autoRecover}
truthAuto <- c(lkin = log(0.7), lkout = log(0.4), ltau = log(1.5), add.sd = 0.3)
cbind(truth = truthAuto,
      posterior = fitAuto$posteriorSummary[names(truthAuto), "mean"],
      `2.5%` = fitAuto$posteriorSummary[names(truthAuto), "2.5%"],
      `97.5%` = fitAuto$posteriorSummary[names(truthAuto), "97.5%"])
```

Two practical notes, both of them about the *solver* rather than about Stan:

- rxode2 detects the delay and picks a dense-output method itself, the same way it does
  under `est = "focei"` or `est = "saem"`; nothing in `stanControl()` mentions DDEs.
- A DDE solve is expensive, and HMC asks for one per gradient.  This example raises
  `max_treedepth` because the default of 10 saturates on it; expect DDE fits to be
  slower per draw than the equivalent ODE fit, and budget iterations accordingly.

## Part 2: the bridge view

The rest of this vignette works the same equation one level lower, through
`rxsRegister()`.  That is the level you want when you are writing the Stan program
yourself -- and it is also the most direct way to *see* the delayed sensitivities, which
the automatic path keeps inside the likelihood.

```{r model}
library(nlmixr2bayes)

dde <- "
kin  <- exp(lkin)
kout <- exp(lkout)
tau  <- exp(ltau)
R(0) <- 1
d/dt(R) <- kin - kout * delay(R, tau)
"
```

The same equation as a bare rxode2 model, with no residual model or priors -- the Stan
program supplies those:

## Register the DDE model

Register it the same way as any other model. All three parameters are differentiated, 
including the delay:

```{r register}
tms <- seq(0.25, 25, by = 0.25)
h <- rxsRegister(dde, events = rxode2::et(tms),
                 sens = c("lkin", "lkout", "ltau"), output = "R",
                 atol = 1e-10, rtol = 1e-10, method = "dop853")
h
```

`method = "dop853"` rather than the default `liblsoda`: the delay machinery needs 
a solver that keeps a dense history to interpolate from.

The handle prints the layout Stan will see -- three parameters in, one state out, one row
per requested time.  That `100 obs -> 100 outputs` line is what the Stan program's
`vector[nObs]` has to match.

## Solve and inspect the sensitivities

`rxsSolve()` calls the same C entry point the Stan program does, so it is the direct way to
see what crosses the bridge:

```{r solve}
truth <- c(kin = 0.7, kout = 0.4, tau = 1.5)
p <- log(unname(truth))
res <- rxsSolve(h, p)
head(res, 3)
```

Column 1 is $R$; the rest are $\partial R/\partial(\ell k_{in}, \ell k_{out}, \ell\tau)$.
The fourth column is exactly zero in those first rows, and that is correct rather than
suspicious: before the first delay has elapsed the solution has not yet seen the lagged
term, so it does not depend on $\tau$ at all.  It becomes non-zero once $t > \tau$:

```{r tauSens}
range(res[tms <= truth[["tau"]], 4])   # before the first delay
range(res[tms >  truth[["tau"]], 4])   # after it
```

## Checking it without trusting finite differences

Before the first delay has elapsed there is a closed form, and it makes a much better 
test than differencing the solver against itself.

While $t \le \tau$ the lagged term is still the constant pre-history, so the equation 
collapses to $R' = k_{in} - k_{out}$, giving $R = 1 + (k_{in} - k_{out})t$. On the 
log scale the sensitivities are then exactly $k_{in}t$ and $-k_{out}t$:

```{r oracle}
early <- which(tms <= truth[["tau"]])
data.frame(t        = tms[early],
           dlkin    = res[early, 2],
           exact    = truth[["kin"]] * tms[early],
           dlkout   = res[early, 3],
           exact2   = -truth[["kout"]] * tms[early])
```

An exact match (to solver tolerances) verifies that the delay itself and its derivatives 
are handled correctly:

```{r oracleMax}
max(abs(res[early, 2] - truth[["kin"]] * tms[early]))
max(abs(res[early, 3] + truth[["kout"]] * tms[early]))
```

Past the first delay interval there is no closed form left, so the fallback there is
central differences of whole solves -- which is what `rxsCheckGradient()` does once the
model is compiled into Stan.

## Using the DDE in a hand-coded Stan program

Once registered, use it in Stan exactly as you would an ODE:

```{r stanCode}
ddeStan <- "
functions {
  vector rx_solve(int handle, vector p);
}
data {
  int<lower=1> nObs;
  int<lower=1> handle;
  vector[nObs] rObs;
}
parameters {
  real lkin;
  real lkout;
  real ltau;
  real<lower=0> sigma;
}
model {
  vector[3] p = [lkin, lkout, ltau]';
  vector[nObs] r = rx_solve(handle, p);

  lkin ~ normal(0, 1);
  lkout ~ normal(-1, 1);
  ltau ~ normal(0, 1);
  sigma ~ exponential(1);

  rObs ~ normal(r, sigma);
}
"
```

Nothing in it mentions a delay: from Stan's side `rx_solve()` is just a function with
known partials.  `rstan::stanc()` confirms it is a valid program -- `allow_undefined`
is what lets `rx_solve()` stay undeclared until the bridge supplies it at link time:

```{r stanc}
sc <- rstan::stanc(model_code = ddeStan, model_name = "dde",
                   allow_undefined = TRUE)
sc$status
```

Compiling and fitting it is then the ordinary pattern from
`vignette("rxstan-handcoded", "nlmixr2bayes")` -- write `ddeStan` to a `.stan` file,
`rxsStanModel()` it, and `rstan::sampling()` with
`list(nObs = length(rObs), handle = h, rObs = rObs)`:

```{r fit, eval = FALSE}
f <- tempfile(fileext = ".stan")
writeLines(ddeStan, f)
sm <- rxsStanModel(f, modelName = "dde")
fit <- rstan::sampling(sm, data = list(nObs = length(rObs), handle = h,
                                       rObs = rObs),
                       chains = 2, iter = 1000)
```

The bridge compiles the rxode2 model with the forward-sensitivity DDE system augmented, 
solves it, and supplies both the solution and its derivatives to Stan's reverse-mode 
autodiff. No DDE solver in Stan; no finite differencing; exact sensitivities.

## Key points

- **DDEs require dense output**: Use `method = "dop853"` or another dense solver, 
  and set `dense = TRUE` in `rxControl()`.
- **The delay itself is a parameter**: You can estimate the lag time $\tau$, and 
  its gradient flows through the sensitivity equations correctly.
- **Pre-history must be constant**: The initial condition `R(0) <- 1` defines both 
  the condition at $t=0$ and the pre-history for $t < 0$. It must be a constant 
  (not a function of time or parameters).
- **The fast path is automatic**: rxstan detects DDEs and enables the fast path
  when safe; `rxsFastAvailable()` reports whether this handle got it (a modelled
  `alag()`/`f()`/`dur()`/`rate()` with dose rows is the case that refuses it):

```{r fast}
rxsFastAvailable(h)
```

Release the handle when you are done with it:

```{r release}
rxsRelease(h)
```

## See also

- `vignette("rxstan-handcoded", "nlmixr2bayes")` -- the general hand-coded
  Stan pattern this vignette specializes for DDEs.
- `vignette("nlmixr2bayes", "nlmixr2bayes")` -- the automatic,
  likelihood-level approach used in the first half of this vignette.
