---
title: "Delay differential equations through rxstan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Delay differential equations through rxstan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Delay differential equations: a Stan limitation

**Stan has no DDE solver.** Not a slow one — none. There is no `ode_*` function that accepts 
a lagged state, and you cannot write one in the Stan language, because the solution has to be 
interpolated at $t - \tau$ while it is still being computed.

However, **rxode2 has `delay()`**, and it propagates sensitivities through the delay correctly. 
With rxstan's hand-coded approach (see `vignette("rxstan-handcoded", "nlmixr2bayes")`),
a DDE becomes an ordinary Stan model.

This vignette walks through a realistic example: delayed indirect response with feedback.

### Other approaches

- **Automatic generation**: `nlmixr2(model, data, est = "stan")` doesn't support DDEs yet. 
  See `vignette("nlmixr2bayes", "nlmixr2bayes")` for automatic estimation of ODE models.
- **Hand-coded ODEs**: See `vignette("rxstan-handcoded", "nlmixr2bayes")` for the general pattern.

---

## A delayed indirect response

Turnover with delayed feedback: the loss term depends on the response one delay ago 
rather than now.

$$\frac{dR}{dt} = k_{in} - k_{out}\,R(t - \tau), \qquad R(s) = 1 \ \text{for}\ s \le 0$$

```{r model, eval=FALSE}
library(nlmixr2)

dde <- "
kin  <- exp(lkin)
kout <- exp(lkout)
tau  <- exp(ltau)
R(0) <- 1
d/dt(R) <- kin - kout * delay(R, tau)
"
```

`R(0) <- 1` does double duty: it is the initial condition *and* the pre-history, 
so `delay(R, tau)` returns 1 whenever $t - \tau \le 0$.

## Register the DDE model

Register it the same way as any other model. All three parameters are differentiated, 
including the delay:

```{r register, eval=FALSE}
tms <- seq(0.25, 25, by = 0.25)
invisible(capture.output(  # swallow rxode2's compile progress bar
  h <- rxsRegister(dde, events = rxode2::et(tms),
                   sens = c("lkin", "lkout", "ltau"), output = "R",
                   atol = 1e-10, rtol = 1e-10, method = "dop853")
))
h
```

`method = "dop853"` rather than the default `liblsoda`: the delay machinery needs 
a solver that keeps a dense history to interpolate from.

## Solve and inspect the sensitivities

```{r solve, eval=FALSE}
truth <- c(kin = 0.7, kout = 0.4, tau = 1.5)
p <- log(unname(truth))
res <- rxsSolve(h, p)
head(res, 3)
```

Column 1 is $R$; the rest are $\partial R/\partial(\ell k_{in}, \ell k_{out}, \ell\tau)$.

## Checking it without trusting finite differences

Before the first delay has elapsed there is a closed form, and it makes a much better 
test than differencing the solver against itself.

While $t \le \tau$ the lagged term is still the constant pre-history, so the equation 
collapses to $R' = k_{in} - k_{out}$, giving $R = 1 + (k_{in} - k_{out})t$. On the 
log scale the sensitivities are then exactly $k_{in}t$ and $-k_{out}t$:

```{r oracle, eval=FALSE}
early <- which(tms <= truth[["tau"]])
data.frame(t        = tms[early],
           dlkin    = res[early, 2],
           exact    = truth[["kin"]] * tms[early],
           dlkout   = res[early, 3],
           exact2   = -truth[["kout"]] * tms[early])
```

An exact match (to solver tolerances) verifies that the delay itself and its derivatives 
are handled correctly.

## Using the DDE in a hand-coded Stan program

Once registered, use it in Stan exactly as you would an ODE:

```stan
functions {
  vector rx_solve(int handle, vector p);
}
data {
  int<lower=1> nObs;
  int<lower=1> handle;
  vector[nObs] tms;
  vector<lower=0>[nObs] rObs;
}
parameters {
  real lkin;
  real lkout;
  real<lower=0> ltau;
  real<lower=0> sigma;
}
model {
  vector[3] p = [lkin, lkout, ltau]';
  vector[nObs] r = rx_solve(handle, p);
  
  lkin ~ normal(0, 1);
  lkout ~ normal(-1, 1);
  ltau ~ normal(0, 1);
  sigma ~ exponential(1);
  
  rObs ~ normal(r, sigma);
}
```

The bridge compiles the rxode2 model with the forward-sensitivity DDE system augmented, 
solves it, and supplies both the solution and its derivatives to Stan's reverse-mode 
autodiff. No DDE solver in Stan; no finite differencing; exact sensitivities.

## Key points

- **DDEs require dense output**: Use `method = "dop853"` or another dense solver, 
  and set `dense = TRUE` in `rxControl()`.
- **The delay itself is a parameter**: You can estimate the lag time $\tau$, and 
  its gradient flows through the sensitivity equations correctly.
- **Pre-history must be constant**: The initial condition `R(0) <- 1` defines both 
  the condition at $t=0$ and the pre-history for $t < 0$. It must be a constant 
  (not a function of time or parameters).
- **The fast path is automatic**: rxstan detects DDEs and enables the fast path
  when safe; use `rxsFastAvailable(handle)` to check.

## See also

- `vignette("rxstan-handcoded", "nlmixr2bayes")` -- the general hand-coded
  Stan pattern this vignette specializes for DDEs.
- `vignette("nlmixr2bayes", "nlmixr2bayes")` -- the automatic,
  likelihood-level approach (does not support DDEs).
