---
title: "nlmixr2bayes and Torsten"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nlmixr2bayes and Torsten}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
## Every port below is generated for real with stanControl(run = FALSE) and
## handed to stanc, which takes seconds -- no sampling, so nothing is cached.
```

[Torsten](https://github.com/metrumresearchgroup/Torsten) (Metrum Research
Group) and nlmixr2bayes both bring Stan's NUTS sampler to pharmacometric
models, from opposite directions:

- **Torsten** is a *library of Stan functions* (`pmx_solve_onecpt`,
  `pmx_solve_twocpt`, `pmx_solve_linode`, `pmx_solve_rk45/bdf/adams`,
  coupled PK-forced-PD solvers) with NONMEM-convention event handling
  (AMT/EVID/CMT/RATE/ADDL/II/SS).  You write the full Stan program by
  hand: data block, parameter transforms, the eta hierarchy, the residual
  model, generated quantities.  Total control, and total responsibility.

- **nlmixr2bayes** is a *full nlmixr2 estimation method*: the model is the
  same `ini({})`/`model({})` function every other nlmixr2 method fits, the
  Stan program is generated, and Stan is **linked** to the compiled
  rxode2/nlmixr2est likelihood (with analytic gradients) rather than the
  model being re-expressed in Stan's language.  The fit comes back as a
  standard `nlmixr2FitData` with GOF tables, plus the full `stanfit`.
  For programs you write yourself, the same package also ships the
  **rxstan** bridge (`vignette("rxstan-handcoded", "nlmixr2bayes")`):
  Torsten-style hand-written Stan, but with rxode2 as the solver instead
  of Stan's own `ode_*` functions or Torsten's `pmx_solve_*`.

The practical consequence: a Torsten model is ~100-200 lines of Stan; the
same model in nlmixr2bayes is the ~20-line nlmixr2 function you may already
have from a FOCEi/SAEM run, and the posterior is sampled against the
*identical* likelihood code those methods used.

## Capability comparison

| capability | Torsten | nlmixr2bayes |
|---|---|---|
| 1/2-cmt analytic models | `pmx_solve_onecpt/twocpt` | `linCmt()` (1/2/3-cmt + absorption) |
| general linear ODE | matrix exponential (`pmx_solve_linode`) | rxode2 solvers |
| general (stiff) ODE | `pmx_solve_rk45/bdf/adams` | rxode2 (lsoda/liblsoda/dop853, ...) |
| delay differential equations | -- | rxode2 `delay()` / `past()` (rxstan hand-coded bridge; the automatic `est="stan"` path does not support DDEs) |
| NONMEM-convention events (AMT/EVID/ADDL/SS/RATE) | yes | yes (rxode2 event table) |
| estimated `alag`/`f`/`dur`/`rate` | parameters of the solve (Torsten notes an open adjoint issue at `tlag = 0`) | yes; jump-condition sensitivities, FD-verified |
| residual models | hand-written (anything Stan can express) | nlmixr2 catalog (add/prop/combined, lnorm, ...), censoring M2/M3/M4, user `ll()` |
| eta structure | hand-written (any structure, incl. IOV by hand) | lotri blocks; non-centered or centered; fixed-variance etas (constant Cholesky factor); first-class IOV (`iov.x ~ v \| OCC` through nlmixr2est's iov hook: per-occasion fixed unit-variance etas scaled by an estimated magnitude theta, FD-verified) |
| finite mixtures | hand-written | `mix()` (2 components): log-sum-exp marginalization, membership posteriors in `fit$env$mixProb` |
| priors | any Stan distribution, hand-written | the 39-distribution lotri catalog via `prior()` in `ini({})`, LKJ/inv-Wishart on omega |
| single-subject (no etas) | yes | yes ("tier 0": the nlm path's likelihood + analytic theta gradients, one external scalar `nlmixr2_pop_ll(theta)`) |
| gradients | Stan autodiff through the solvers | analytic forward sensitivities via `precomputed_gradients` |
| within-chain parallelism | MPI (`map_rect`-style population solvers) | subject-parallel OpenMP inside each likelihood evaluation (`cores`) |
| interface | cmdstan / cmdstanr | rstan, `nlmixr2(..., est = "stan")` |
| output | cmdstan CSV; roll your own diagnostics/GOF | `nlmixr2FitData` (IPRED/CWRES tables, `$objDf` with FOCEi/WAIC/LOO rows) + the `stanfit` |
| model reuse across estimation methods | Stan-only | the same function fits with focei/saem/bobyqa/... |

Where Torsten is the right tool: hand-tuned Stan programs, structures
outside the nlmixr2 ui (e.g. exotic hierarchies), and cmdstan-ecosystem
workflows.  Where nlmixr2bayes is the right tool: standard
population PK/PD where the model already exists in nlmixr2 form, and the
Bayesian run should be comparable line-for-line with the frequentist ones.

There is also a middle ground: an **rxode2 sensitivity linkage** for
hand-written Stan programs (the rxstan bridge covered in
`vignette("rxstan-handcoded", "nlmixr2bayes")`) -- write your own Stan
code (any hierarchy, any prior structure) while the ODE system, dosing
events, and their analytic sensitivities come from a compiled rxode2
model (including event-time "jump" sensitivities and delay differential
equations, which no Stan-native solver provides).  It shares the same
no-Stan-code-in-the-linked-`.so` architecture as the automatic path,
just at the ODE level instead of the likelihood level.

## How the ports below are checked

Each port is a real nlmixr2 model, and each one is put through `est = "stan"`
code generation with `run = FALSE` -- which builds the complete Stan program
and its data list without touching the sampler -- and the result is handed to
`rstan::stanc()`.  A program that stanc accepts is a program that would have
compiled and sampled; that is what keeps these ports honest without needing
cmdstan and Torsten installed.  (The same checks run in the package's own
`test-torsten-port.R`.)

```{r helper}
library(nlmixr2)
library(nlmixr2bayes)

## generate the Stan program for a model + data, and assert stanc accepts it
portCheck <- function(model, data) {
  code <- nlmixr2est::nlmixr2(model, data, est = "stan",
                              control = stanControl(run = FALSE))
  stopifnot(rstan::stanc(model_code = code$code,
                         allow_undefined = TRUE)$status)
  code
}
```

The datasets below are small synthetic event tables in the NONMEM convention,
enough to exercise generation.  Sampling-level agreement with Torsten itself
would need cmdstan plus Torsten's own example data; see "A note on
completeness" at the end.

## Port: `poppk2cpt` (two-compartment population model)

Torsten's `example-models/poppk2cpt` is ~150 lines of Stan:
NONMEM-style arrays in `data`, `pmx_solve_twocpt` per subject, Matt's
trick (Matt Hoffman) for the etas, lognormal priors, lognormal
residual.  The same model in nlmixr2bayes:

```{r poppk2cpt}
poppk2cpt <- function() {
  ini({
    lcl <- log(10)
    lq <- log(15)
    lv1 <- log(35)
    lv2 <- log(105)
    lka <- log(2.5)
    # Torsten's lognormal priors on the natural scale are normal priors on
    # the log-scale parameters
    prior(lcl) ~ dnorm(log(10), 0.25)
    prior(lq) ~ dnorm(log(15), 0.5)
    prior(lv1) ~ dnorm(log(35), 0.25)
    prior(lv2) ~ dnorm(log(105), 0.5)
    prior(lka) ~ dnorm(log(2.5), 1)
    lnorm.sd <- c(0, 0.2)
    prior(lnorm.sd) ~ dcauchy(0, 1)
    # IIV block; the default omega prior is announced as
    # lkj_corr_cholesky(2) + half-Cauchy SDs -- the same shape Torsten's
    # examples hand-code
    eta.cl + eta.q + eta.v1 + eta.v2 + eta.ka ~
      c(0.06,
        0.01, 0.06,
        0.01, 0.01, 0.06,
        0.01, 0.01, 0.01, 0.06,
        0.01, 0.01, 0.01, 0.01, 0.06)
  })
  model({
    cl <- exp(lcl + eta.cl)
    q <- exp(lq + eta.q)
    v <- exp(lv1 + eta.v1)    # central volume (Torsten's V1)
    vp <- exp(lv2 + eta.v2)   # peripheral volume (Torsten's V2)
    ka <- exp(lka + eta.ka)
    # pmx_solve_twocpt's analytic solution == linCmt() with these params
    cp <- linCmt()
    # logCObs ~ normal(log(cHat), sigma) == a lognormal residual
    cp ~ lnorm(lnorm.sd)
  })
}
```

The event table is the same NONMEM-convention data both tools consume --
no `iObs`/`start`/`end` index bookkeeping; rxode2 derives it:

```{r poppk2cptData}
d <- do.call(rbind, lapply(1:4, function(id) {
  rbind(data.frame(ID = id, TIME = 0, DV = NA_real_, AMT = 1000, EVID = 1),
        data.frame(ID = id, TIME = c(1, 2, 4, 8, 12, 24),
                   DV = c(20, 25, 18, 10, 6, 2), AMT = 0, EVID = 0))
}))
codePop <- portCheck(poppk2cpt, d)
```

A real run is then `nlmixr2(poppk2cpt, d, est = "nuts")`.  The generated
program is what it would sample; the correlated five-eta block is where
Torsten's hand-coded Matt's trick shows up, emitted rather than written:

```{r poppk2cptCode}
cat(grep("Lcorr|lkj|^\\s*z\\b|eta\\[|matrix", strsplit(codePop$code, "\n")[[1]],
         value = TRUE), sep = "\n")
```

A note on the eta parameterization: nlmixr2bayes's generated program uses
**the same Matt's trick** Torsten's examples hand-code -- by default the
etas are non-centered, with `z ~ std_normal()` sampled and
`eta = z * L'` assembled from the omega Cholesky factor in
`transformed parameters` -- it is just emitted for you rather than written
by hand.  `stanControl(etaParam = "centered")` switches to the centered
form (`eta[i] ~ multi_normal_cholesky(0, L)`); the two are verified to
target the identical posterior.

```{r centered}
codeCentered <- nlmixr2est::nlmixr2(
  poppk2cpt, d, est = "stan",
  control = stanControl(run = FALSE, etaParam = "centered"))
cat(grep("multi_normal_cholesky|std_normal",
         strsplit(codeCentered$code, "\n")[[1]], value = TRUE), sep = "\n")
```

## Port: Friberg-Karlsson neutropenia (population PK/PD, two endpoints)

Torsten's `FribergKarlsson` example: a two-compartment PK driving the
classic proliferation/transit/circulation myelosuppression chain, ~200
lines with a hand-written 8-state ODE function and separate PK/PD
observation indexing.  In nlmixr2 syntax:

```{r fk}
fk <- function() {
  ini({
    lcl <- log(10); lq <- log(15); lv1 <- log(35); lv2 <- log(105)
    lka <- log(2)
    lmtt <- log(125)
    lcirc0 <- log(5)
    lalpha <- log(3e-4)
    lgamma <- log(0.17)
    prior(lcl) ~ dnorm(log(10), 0.25)
    prior(lmtt) ~ dnorm(log(125), 0.25)
    prior(lcirc0) ~ dnorm(log(5), 0.25)
    prior(lalpha) ~ dnorm(log(3e-4), 1)
    prior(lgamma) ~ dnorm(log(0.17), 0.5)
    lnorm.sd <- c(0, 0.2)
    lnormNeut.sd <- c(0, 0.3)
    prior(lnorm.sd) ~ dcauchy(0, 1)
    prior(lnormNeut.sd) ~ dcauchy(0, 1)
    eta.cl ~ 0.06
    eta.mtt ~ 0.04
    eta.circ0 ~ 0.04
  })
  model({
    cl <- exp(lcl + eta.cl)
    q <- exp(lq); v1 <- exp(lv1); v2 <- exp(lv2); ka <- exp(lka)
    mtt <- exp(lmtt + eta.mtt)
    circ0 <- exp(lcirc0 + eta.circ0)
    alpha <- exp(lalpha)
    gamma <- exp(lgamma)
    k10 <- cl / v1; k12 <- q / v1; k21 <- q / v2
    ktr <- 4 / mtt
    d/dt(depot) <- -ka * depot
    d/dt(cent) <- ka * depot - (k10 + k12) * cent + k21 * peri
    d/dt(peri) <- k12 * cent - k21 * peri
    conc <- cent / v1
    edrug <- alpha * conc
    d/dt(prol) <- ktr * prol * (1 - edrug) * (circ0 / circ)^gamma - ktr * prol
    d/dt(tr1) <- ktr * (prol - tr1)
    d/dt(tr2) <- ktr * (tr1 - tr2)
    d/dt(tr3) <- ktr * (tr2 - tr3)
    d/dt(circ) <- ktr * tr3 - ktr * circ
    prol(0) <- circ0; tr1(0) <- circ0; tr2(0) <- circ0; tr3(0) <- circ0
    circ(0) <- circ0
    # two endpoints, distinguished by CMT/DVID in the data -- Torsten does
    # this with separate iObsPK/iObsPD index arrays
    conc ~ lnorm(lnorm.sd)
    circ ~ lnorm(lnormNeut.sd)
  })
}
```

The two endpoints are told apart by `CMT`, and nothing else about the data
changes:

```{r fkData}
dFk <- do.call(rbind, lapply(1:3, function(id) {
  rbind(data.frame(ID = id, TIME = 0, DV = NA_real_, AMT = 80, EVID = 1,
                   CMT = "depot"),
        data.frame(ID = id, TIME = c(1, 4, 12, 24),
                   DV = c(1.5, 1.1, 0.5, 0.2), AMT = 0, EVID = 0,
                   CMT = "conc"),
        data.frame(ID = id, TIME = c(24, 96, 168, 336),
                   DV = c(4.9, 3.2, 2.1, 4.4), AMT = 0, EVID = 0,
                   CMT = "circ"))
}))
codeFk <- portCheck(fk, dFk)
```

Multi-endpoint likelihoods, initial conditions, and the stiff ODE solve are
all standard rxode2/nlmixr2est machinery; `likInner0` sums the per-endpoint
terms and the analytic gradients flow through the same linkage.  None of that
shows up in the Stan program, which is the point -- the whole 8-state system,
both residual models and both endpoints reduce to one external call:

```{r fkCode}
cat(grep("nlmixr2_cond_all2|^\\s*(vector|matrix|real|array)",
         strsplit(codeFk$code, "\n")[[1]], value = TRUE)[1:12], sep = "\n")
```

## Port: `effCpt` (effect-compartment population PK/PD)

Torsten's `effCpt` example: a two-compartment oral PK linked to an effect
compartment (`ke0`) driving an Emax response, fit to concentration and
response observations simultaneously (~250 lines of Stan with a
hand-built rate matrix and separate PK/PD indexing):

```{r effCpt}
effCpt <- function() {
  ini({
    lcl <- log(10); lq <- log(15); lv1 <- log(35); lv2 <- log(105)
    lka <- log(2)
    lke0 <- log(1)
    lec50 <- log(100)
    prior(lcl) ~ dnorm(log(10), 0.25)
    prior(lke0) ~ dnorm(log(1), 0.25)
    prior(lec50) ~ dnorm(log(100), 0.5)
    lnorm.sd <- c(0, 0.2)
    resp.sd <- c(0, 5)
    prior(lnorm.sd) ~ dcauchy(0, 1)
    prior(resp.sd) ~ dcauchy(0, 5)
    eta.cl ~ 0.06
    eta.v1 ~ 0.06
    eta.ke0 ~ 0.04
    eta.ec50 ~ 0.04
  })
  model({
    cl <- exp(lcl + eta.cl)
    q <- exp(lq)
    v1 <- exp(lv1 + eta.v1)
    v2 <- exp(lv2)
    ka <- exp(lka)
    ke0 <- exp(lke0 + eta.ke0)
    ec50 <- exp(lec50 + eta.ec50)
    k10 <- cl / v1
    k12 <- q / v1
    k21 <- q / v2
    d/dt(depot) <- -ka * depot
    d/dt(cent) <- ka * depot - (k10 + k12) * cent + k21 * peri
    d/dt(peri) <- k12 * cent - k21 * peri
    cp <- 1000 * cent / v1
    d/dt(ce) <- ke0 * (cp - ce)          # effect compartment
    resp <- 100 * ce / (ec50 + ce)       # Emax(100) response
    cp ~ lnorm(lnorm.sd)
    resp ~ add(resp.sd)
  })
}

dEff <- do.call(rbind, lapply(1:3, function(id) {
  rbind(data.frame(ID = id, TIME = 0, DV = NA_real_, AMT = 100, EVID = 1,
                   CMT = "depot"),
        data.frame(ID = id, TIME = c(0.5, 1, 2, 4, 8, 12),
                   DV = c(15, 22, 24, 18, 12, 8), AMT = 0, EVID = 0,
                   CMT = "cp"),
        data.frame(ID = id, TIME = c(1, 2, 4, 8, 12, 24),
                   DV = c(10, 25, 40, 35, 25, 12), AMT = 0, EVID = 0,
                   CMT = "resp"))
}))
codeEff <- portCheck(effCpt, dEff)
```

Both endpoints ride the same solve; `likInner0` sums the per-endpoint
terms, and the analytic gradients flow through the same linkage.  Note that
the two residual models are *different* families (lognormal on concentration,
additive on response) and both are ordinary `ini({})` parameters with their
own priors:

```{r effCptCode}
cat(grep("lnorm_sd|resp_sd", strsplit(codeEff$code, "\n")[[1]], value = TRUE),
    sep = "\n")
```

## IOV

Torsten "supports" IOV the way it supports everything: you write it --
occasion-level etas declared as extra parameters, constrained by hand to
share one variance, and added into the occasion's parameters.
nlmixr2bayes has first-class syntax for the same structure -- `iov.x ~ v
| occ` declares one shared IOV magnitude across occasions:

```{r iov}
iov <- function() {
  ini({
    lcl <- log(10)
    lv <- log(35)
    prior(lcl) ~ dnorm(log(10), 0.5)
    lnorm.sd <- c(0, 0.2)
    prior(lnorm.sd) ~ dcauchy(0, 1)
    eta.cl ~ 0.06       # IIV
    iov.cl ~ 0.02 | OCC # IOV: one shared magnitude across occasions
  })
  model({
    cl <- exp(lcl + eta.cl + iov.cl)
    v <- exp(lv)
    cp <- linCmt()
    cp ~ lnorm(lnorm.sd)
  })
}

dIov <- do.call(rbind, lapply(1:4, function(id) {
  do.call(rbind, lapply(1:2, function(occ) {
    t0 <- (occ - 1) * 24
    rbind(data.frame(ID = id, TIME = t0, DV = NA_real_, AMT = 1000, EVID = 1,
                     OCC = occ),
          data.frame(ID = id, TIME = t0 + c(1, 2, 4, 8, 12),
                     DV = c(20, 25, 18, 10, 6), AMT = 0, EVID = 0, OCC = occ))
  }))
}))
codeIov <- portCheck(iov, dIov)
```

Under the hood, nlmixr2est's preprocessing hook expands `iov.cl` into
per-occasion etas with variance **fixed to one**, and a single estimated
`sd` theta carries the IOV magnitude -- the same construction a Torsten
user would otherwise hand-write (a unit-variance parameter scaled by a
shared SD), except emitted for you and FD-verified end to end.  The
expansion is visible in the generated parameter block, where the IOV
magnitude appears as an ordinary theta:

```{r iovCode}
cat(grep("iov", strsplit(codeIov$code, "\n")[[1]], value = TRUE,
         ignore.case = TRUE), sep = "\n")
```

The linked likelihood never sees anything but ordinary subject-level etas: the
expansion happens before the Stan program is generated, so no upstream
likelihood change was needed to support it.

## What does not carry over

- Torsten's **single-patient** examples (`pk2cpt` and friends) carry over
  directly: a no-eta model generates a theta-only program with
  `target += nlmixr2_pop_ll(theta)` (nlmixr2est's nlm path -- the
  likelihood and analytic theta gradients on the natural scale).  The
  pk2cpt port is the population port above minus the etas, with Torsten's
  lognormal priors verbatim:

```{r pk2cpt}
pk2cpt <- function() {
  ini({
    lcl <- log(10); lq <- log(15); lv1 <- log(35); lv2 <- log(105)
    lka <- log(2.5)
    prior(lcl) ~ dnorm(log(10), 0.25)
    prior(lq) ~ dnorm(log(15), 0.5)
    prior(lv1) ~ dnorm(log(35), 0.25)
    prior(lv2) ~ dnorm(log(105), 0.5)
    prior(lka) ~ dnorm(log(2.5), 1)
    lnorm.sd <- c(0, 0.2)
    prior(lnorm.sd) ~ dcauchy(0, 1)
  })
  model({
    cl <- exp(lcl); q <- exp(lq); v <- exp(lv1); vp <- exp(lv2)
    ka <- exp(lka)
    cp <- linCmt()
    cp ~ lnorm(lnorm.sd)
  })
}

d1 <- rbind(data.frame(ID = 1, TIME = 0, DV = NA_real_, AMT = 1000, EVID = 1),
            data.frame(ID = 1, TIME = c(0.5, 1, 2, 4, 8, 12, 24),
                       DV = c(15, 22, 24, 18, 12, 8, 3), AMT = 0, EVID = 0))
code1 <- portCheck(pk2cpt, d1)
cat(grep("nlmixr2_pop_ll", strsplit(code1$code, "\n")[[1]], value = TRUE),
    sep = "\n")
```

  That single external scalar is the whole difference: with no etas there is
  nothing to marginalize, so nlmixr2est's nlm path hands Stan the complete
  data log-likelihood and its analytic theta gradient in one call.  The
  tier-0 target is verified against a hand-written density with FD-checked
  gradients.

- Torsten's `pmx_solve_group_*` **MPI parallelism** distributes subjects
  across processes; nlmixr2bayes parallelizes subjects with OpenMP threads
  inside each likelihood call (`cores`) -- same axis, different
  mechanics, no user-visible model difference.

- Conversely, rxode2 capabilities with no Torsten equivalent flow through
  `est="stan"` unchanged: the censoring machinery (M2/M3/M4), the
  residual-error catalog, transit-compartment helpers, and
  mu-referenced covariate handling.  Delay differential equations are the
  one rxode2 capability Torsten also lacks but that nlmixr2bayes does not
  yet expose automatically -- reach them through the rxstan hand-coded
  bridge (`vignette("rxstan-dde", "nlmixr2bayes")`) instead.

## A note on completeness

These ports are verified at the code level -- each generates a program stanc
accepts, which is exactly what the chunks above did -- but they have **not**
been run against Torsten's own example datasets, which ship inside the Torsten
repository as R dump files rather than in any R package.  The synthetic event
tables above exercise generation, not agreement.

To reproduce a port against the original data, load Torsten's dump and reshape
the NONMEM-style vectors into the event-table columns nlmixr2 expects:

```{r torstenData, eval = FALSE}
e <- new.env()
source(file.path(torstenDir, "example-models/poppk2cpt/poppk2cpt.data.R"),
       local = e)
d <- with(e, data.frame(ID = rep(seq_len(nSubjects),
                                 times = end - start + 1),
                        TIME = time, AMT = amt, EVID = evid, CMT = cmt,
                        RATE = rate, II = ii, ADDL = addl, SS = ss,
                        DV = NA_real_))
d$DV[e$iObs] <- e$cObs
fit <- nlmixr2est::nlmixr2(poppk2cpt, d, est = "nuts",
                           control = nutsControl(seed = 42))
```

That chunk is the only one on this page that does not run when the vignette is
built -- `torstenDir` has to point at a clone of the Torsten repository.
